Conflict-Free Vertex Coloring Of Planar Graphs

### 2017-05-01 paper/paper.tex
Last edits to paper before final copy

\maketitle
\begin{abstract}
The conflict-free coloring problem is a variation of the vertex coloring problem, a classical NP-hard optimization problem. The conflict-free coloring problem aims to color a possibly proper subset of vertices such that there is a unique color within the closed neighborhood (a vertex and its neighbors) of every vertex. This paper presents recent heuristics to solve the conflict-free coloring problem on both general graphs and planar graphs.
\end{abstract}

@@

A simple graph is a loopless graph where no two edges connect the same pair of vertices. A planar graph is a graph that can be drawn on the plane such that its edges only intersect at their endpoints. This drawing is referred to as a planar drawing. All graphs used by the vertex coloring problem and conflict-free coloring problem are assumed to be simple graphs. An example simple graph that we will use later on is shown in Figure \ref{fig:graph}.

The neighborhood of a vertex, denoted $N_G(v)$, is a set of all vertices adjacent to $v$. A \emph{closed} neighborhood consists the vertices adjacent to $v$ plus $v$ itself. For example, let's examine $N_{G_1}(1)$ from Figure \ref{fig:graph}. The open neighborhood of 1 is $\{2, 5\}$ while the closed neighborhood is $\{1, 2, 5\}$. For the rest of the paper, the term \emph{neighborhood} will refer to closed neighborhoods unless stated otherwise.

@@

A graph $G$ is said to be \emph{k-colorable} if it can be colored using $k$ or fewer colors, i.e. $\chi(G) \leq k$. A graph having $\chi(G) = k$ is said to be a \emph{k-chromatic} graph. The \emph{k-colorability} problem asks if a graph can be colored using $k$ colors. This problem is slightly easier than the VCP as it is decision problem rather than an optimization problem. An example of a 3-colorable graph and a vertex coloring is shown in Figure \ref{fig:vcp-example}.

@@

An example of a conflict-free coloring is shown in Figure \ref{fig:cfcp-example}. All uncolored vertices are in gray. It is a valid conflict-free coloring because every vertex has a unique color within its neighborhood. For example, $N(1) = \{1: uncolored,\ 2: uncolored,\ 5: red\}$. Thus, red is the unique color within the neighborhood of vertex 1. We observe from Figures \ref{fig:vcp-example} and \ref{fig:cfcp-example}, $\chi(G_1) = 3$ and $\chi_{CF}(G_1) = 2$. It is worth noting that all proper vertex colorings are also conflict-free colorings. This is because in a proper vertex coloring, every vertex is its own conflict-free neighbor \cite{abel2017three}.

@@

The k-colorability problem as well as the conflict-free k-colorability problem are NP-complete. This can be proven by using a reduction from a known NP-complete problem. If we find a known hard problem $Y$, we can prove that another problem $X$ is hard by a reduction from $Y$ to $X$. The VCP and the CFCP have both been shown to be NP-hard \cite{abel2017three,moret1998theory}. The \emph{k-colorability} problem is proven to be NP-complete with a reduction from 3-SAT, a well-known NP-complete problem \cite{sharma2012new}.

@@

\subsection{Guaranteeing CF k-Colorability}
It is wasteful to spend time coloring a graph that cannot be conflict-free colored. This leads Abel et al. to provide sufficient criterion to guarantee the conflict-free k-colorability of a certain graphs. To introduce this criterion, we need to introduce a few specific graphs. A complete graph is a simple, undirected graph where every pair of distinct vertices is connected by an edge. A complete graph on $n$ vertices is denoted as $K_n$. The graph $K_n^{-3}$ is the graph obtained by removing any three edges forming a single triangle, i.e. $K_3$.

@@

Abel et al. give a criterion that guarantees k-colorability which is stated in Theorem \ref{thm:criterion}. In the proof of this theorem, found in \cite{abel2017three}, the authors prove that their heuristic will always generate a k-colorable graph when the specified criterion is met. To illustrate this heuristic, we will use a graph that fits the specified criterion and apply the iterated elimination of distance-3-sets algorithm defined in Algorithm \ref{alg:elimination}.

@@

We will be focusing on the simplest example, $k = 1$. Graph $G_2$ is a simple graph with 6 vertices. It does not, however, fit our criterion; it actually breaks both properties as it has both $K_{k+2}$ and $K_{k+4}^{-3}$ as a minor. For $k = 1$, this means our graph cannot contain $K_{3}$ or $K_{4}^{-3}$ as a minor. These graphs are shown in Figure \ref{fig:bad-criterion}.

@@

We can see that $G_2$ has both of these as minors. The transformation needed to see the minors is shown in Figure \ref{fig:criterion-minor}. By contracting the dotted edges, vertices 1 and 4 would combine as well as vertices 3 and 5. We would have both $K_{3}$ and $K_{4}^{-3}$. Thus, this graph cannot be conflict-free colored with one color.

@@

\vspace{-0.5cm}

\subsection{Iterated Elimination of Distance-3-Sets}
\label{sec:iterated}
The iterated elimination of distance-3-sets algorithm, defined in \cite{abel2017three}, is a polynomial-time heuristic for generating a conflict-free k-coloring of general graphs that satisfy the criterion given in Theorem \ref{thm:criterion}. If the criterion is not met, the heuristic will not generate a valid conflict-free coloring of $k$ colors.

@@

Next, for every vertex in $D$, we must remove their closed neighborhood from the graph. The neighborhood of vertex 2 consists of $\{1,2,3,6\}$ and the neighborhood of vertex 8 consists of $\{4,8\}$. By removing these vertices and any edges containing these vertices, we are left with vertices $\{5, 7\}$. This is shown in Figure \ref{fig:algorithm1-step2}.

@@

\subsection{Bounds for Planar Graphs}
\label{sec:bounds}
Recall the map of the USA example given in the introduction. This is an example of a planar graph and illustrates a famous theorem associated with it. We present this theorem, called the four color theorem, in Theorem \ref{thm:four}. The proof by computer that is widely accepted today can be found in \cite{robertson1997four}.

@@

\subsection{CF Coloring via Dominating Set}
\label{sec:coloring-dom}
Algorithm \ref{alg:elimination} presented in section \ref{sec:iterated} gives us an efficient heuristic for coloring general graphs including planar graphs. Algorithm \ref{alg:cf-dominating}, called conflict-free via dominating set, gives us another way of coloring planar graphs, albeit likely less efficient.

@@

\section{Conclusion}
\label{sec:conclusion}
I have presented some of the recent heuristics and findings on the conflict-free coloring problem. Although the CFCP is NP-hard and finding the minimum number colors needed requires inefficient brute force algorithms, there are heuristics to find good conflict-free colorings.

@@

There is also recent interest in variations of the CFCP such as requiring \emph{another} vertex (not the selected vertex) to be colored within the neighborhood of a selected vertex. This allows for applications such as guiding a specific robot to other locations (i.e. the robot and its destination would both be a distinct unique color).

>>>>

\maketitle
\begin{abstract}
The conflict-free coloring problem is a variation of the vertex coloring problem, a classical NP-hard optimization problem. The conflict-free coloring problem aims to color a possibly proper subset of vertices such that there is a unique color within the closed neighborhood (a vertex and its neighbors) of every vertex. This paper presents recent findings and heuristics to solve the conflict-free coloring problem on both general graphs and planar graphs.
\end{abstract}

@@

A simple graph is a loopless graph where no two edges connect the same pair of vertices. A planar graph is a graph that can be drawn on the plane such that its edges only intersect at their endpoints. This drawing is referred to as a planar drawing. All graphs used by the vertex coloring problem and conflict-free coloring problem are assumed to be simple graphs. An example graph is shown in Figure \ref{fig:graph}.

The neighborhood of a vertex, denoted $N_G(v)$, is a set of all vertices adjacent to $v$. A \emph{closed} neighborhood consists of the vertices adjacent to $v$ plus $v$ itself. For example, let's examine $N_{G_1}(1)$ from Figure \ref{fig:graph}. The open neighborhood of 1 is $\{2, 5\}$ while the closed neighborhood is $\{1, 2, 5\}$. For the rest of the paper, the term \emph{neighborhood} will refer to closed neighborhoods unless stated otherwise.

@@

A graph $G$ is said to be \emph{k-colorable} if it can be colored using $k$ or fewer colors, i.e. $\chi(G) \leq k$. A graph having $\chi(G) = k$ is said to be a \emph{k-chromatic} graph. The \emph{k-colorability} problem asks if a graph can be colored using $k$ colors. This problem is slightly different than the VCP as it is decision problem rather than an optimization problem. An example of a 3-colorable graph and a vertex coloring is shown in Figure \ref{fig:vcp-example}.

@@

An example of a conflict-free coloring is shown in Figure \ref{fig:cfcp-example}. All uncolored vertices are in gray. It is a valid conflict-free coloring because every vertex has a unique color within its neighborhood. For example, $N(1) = \{1: uncolored,\ 2: uncolored,\ 5: red\}$. As there are no other vertices colored red, we have verified there is a unique color in the neighborhood of vertex 1. We observe from Figures \ref{fig:vcp-example} and \ref{fig:cfcp-example}, $\chi(G_1) = 3$ and $\chi_{CF}(G_1) = 2$. It is worth noting that all proper vertex colorings are also conflict-free colorings. This is because in a proper vertex coloring, every vertex is its own conflict-free neighbor \cite{abel2017three}.

@@

The k-colorability problem as well as the conflict-free k-colorability problem are NP-complete. This can be proven by using a reduction from a known NP-complete problem \cite{garey2002computers}. If we find a known hard problem $Y$, we can prove that another problem $X$ is hard by a reduction from $Y$ to $X$. The VCP and the CFCP have both been shown to be NP-hard \cite{abel2017three,moret1998theory}. The \emph{k-colorability} problem is proven to be NP-complete with a reduction from 3-SAT, another well-known NP-complete problem \cite{sharma2012new}. The \emph{conflict-free k-colorability problem} is shown to be NP-complete by a reduction from k-colorability \cite{abel2017three}.

@@

\subsection{Guaranteeing CF k-Colorability}
It is wasteful to spend time coloring a graph that cannot be conflict-free colored. This leads Abel et al. to provide sufficient criterion to guarantee the conflict-free k-colorability of a certain graphs. To introduce this criterion, we need to introduce a few specific graphs. A complete graph is a simple, undirected graph where every pair of distinct vertices is connected by an edge. A complete graph on $n$ vertices is denoted as $K_n$. The graph $K_n^{-3}$ is the graph obtained by removing any three edges forming one single triangle, i.e. $K_3$.

@@

Abel et al. give a criterion that guarantees k-colorability which is stated in Theorem \ref{thm:criterion}. In the proof of this theorem, found in \cite{abel2017three}, the authors prove that their heuristic, given in section \ref{sec:iterated}, will always generate a k-colorable graph when the specified criterion is met.

@@

We will be focusing on the simplest example, $k = 1$. Graph $G_2$ is a simple graph with 6 vertices. For $k = 1$, this means our graph cannot contain $K_{3}$ or $K_{4}^{-3}$ as a minor. These graphs are shown in Figure \ref{fig:bad-criterion}.

@@

We can see that $G_2$ has both of these as minors. The transformation needed to see the minors is shown in Figure \ref{fig:criterion-minor}. By contracting the dotted edges, vertices 1 and 4 would combine as well as vertices 3 and 5. We would have both $K_{3}$ and $K_{4}^{-3}$. This graph does not fit our criterion; it even breaks both properties as it has both $K_{k+2}$ and $K_{k+4}^{-3}$ as a minor. Thus, this graph cannot be conflict-free colored with one color.

@@

\vspace{-0.1cm}

\subsection{Iterated Elimination of Distance-3-Sets}
\label{sec:iterated}
The iterated elimination of distance-3-sets algorithm, defined in \cite{abel2017three} and shown in Algorithm \ref{alg:elimination}, is a polynomial-time heuristic for generating a conflict-free k-coloring of general graphs that satisfy the criterion given in Theorem \ref{thm:criterion}. If the criterion is not met, the heuristic will produce a conflict-free coloring of more than $k$ colors.

@@

Next, for every vertex in $D$, we must remove their closed neighborhood from the graph. The neighborhood of vertex 2 consists of $\{1,2,3,6\}$ and the neighborhood of vertex 8 consists of $\{4,8\}$. By removing these vertices and any edges containing these vertices, we are left with vertices $\{5, 7\}$.~This is shown in Figure \ref{fig:algorithm1-step2}.

@@

\subsection{Bounds for Planar Graphs}
\label{sec:bounds}
Recall the map of the USA example given in the introduction. This is an example of a planar graph and illustrates a famous theorem associated with it. We present this theorem, called the four color theorem, in Theorem \ref{thm:four}. The most widely-accepted proof, which utilizes a computer to brute force every possible case, can be found in \cite{robertson1997four}.

@@

\subsection{CF Coloring via Dominating Set}
\label{sec:coloring-dom}
Algorithm \ref{alg:elimination} presented in section \ref{sec:iterated} gives us an efficient heuristic for coloring general graphs including planar graphs. Algorithm \ref{alg:cf-dominating}, called conflict-free via dominating set, gives us another way of coloring planar graphs, albeit likely less efficient with regards to running time.

@@

\vspace{-0.1cm}

\section{Conclusion}
\label{sec:conclusion}
We have presented some of the recent heuristics and findings on the conflict-free coloring problem. Although the CFCP is NP-hard and finding the minimum number colors needed requires inefficient brute force algorithms, there are heuristics to find good conflict-free colorings.

@@

There is also recent interest in variations of the CFCP such as requiring \emph{another} vertex (not the selected vertex) to be colored within the neighborhood of a selected vertex. This allows for applications such as guiding a specific robot to other locations (i.e. the robot and its destination would both be a distinct unique color). Overall, finding properties and heuristics for conflict-free coloring is beneficial and required for efficiently solving real-world problems.

\vspace{-0.1cm}
